---
title: "R Notebook"
output: html_notebook
---

```{r, echo=FALSE, warning=FALSE}
library(ggplot2)
library(knitr)
library(dplyr)
```


```{r}
newstats <- read.csv("mlb_3season_clean.csv")
allstats <- read.csv("allplayers_3season.csv")
```

```{r}
newstats <- newstats[newstats$year == "2024",]

```


```{r}
colnames(newstats)[4] <- "age"
colnames(allstats)[4] <- "age"
```


```{r}
agestats <- allstats %>%
  group_by(age) %>%
  summarise(total_ab=sum(ab),
             total_hits=sum(hit))


agestats$ageBA <- agestats$total_hits / agestats$total_ab
agestats$norm_hits <- log(agestats$total_hits)
agestats$norm_ba <- log(agestats$ageBA)
agestats$norm_hits_x_ba <- agestats$norm_hits * agestats$ageBA
```

```{r}
joinedstats <- left_join(newstats, agestats, by = "age")
#joinedstats$k_pct <- (joinedstats$strikeout / joinedstats$pa) * 100

joinedstats = subset(joinedstats, select = -c(total_ab,total_hits, norm_hits, norm_ba) )

joinedstats$sb_pa <- joinedstats$r_total_stolen_base / joinedstats$pa
```


```{r}
normage <- agestats %>%
  arrange(desc(age))

normage = subset(normage, select = -c(norm_ba) )


kable(normage, caption = "Every Age Combined Statistics 2022-2023")
```


```{r}
library(mgcv)
mod <- lm(hit ~ ab + launch_angle_avg + sweet_spot_percent + k_percent + sprint_speed + avg_hyper_speed, data=joinedstats)

summary(mod)
```


```{r}
predictions <- predict(mod, joinedstats)
joinedstats$xHits <- predictions

# Find the difference between expected hits and true hits
joinedstats$hit_diff <- joinedstats$xHits - joinedstats$hit

```


```{r, warning=FALSE}
library(dplyr)
joinedstats %>%
  ggplot(aes(xHits, hit)) +
  geom_point() +
  geom_smooth() +
  ggtitle('Predicted Hits vs. Actual')

```


```{r}
reducedjoin <- joinedstats[, c("last_name", "first_name", "age", "ab", "pa", "hit", "batting_avg", "xHits", "hit_diff", "r_total_stolen_base", "norm_hits_x_ba", "ageBA")]

```


Since the number of at-bats varies for every player, I will divide the difference from expected hits and actual hits by the total number of at-bats this past season. From there, I will normalize the values on a 0 to 1 scale. The values after applying the division are quite small, and this will allow the analysis to go much smoother. The equation will look like the following:
$y = norm(\frac{xH - H}{AB})$

```{r}
reducedjoin$hitDiffperAB <- reducedjoin$hit_diff / reducedjoin$ab


min <- min(reducedjoin$hitDiffperAB)
max <- max(reducedjoin$hitDiffperAB)

reducedjoin$normDiffperAB <- (reducedjoin$hitDiffperAB - min)/(max - min)
```


```{r, echo=FALSE}
names(reducedjoin)[names(reducedjoin) == "last_name"] <- "Last"
names(reducedjoin)[names(reducedjoin) == "first_name"] <- "First"
names(reducedjoin)[names(reducedjoin) == "age"] <- "Age"
names(reducedjoin)[names(reducedjoin) == "ab"] <- "AB"
names(reducedjoin)[names(reducedjoin) == "r_total_stolen_base"] <- "SB"
names(reducedjoin)[names(reducedjoin) == "pa"] <- "PA"
names(reducedjoin)[names(reducedjoin) == "hit"] <- "H"
names(reducedjoin)[names(reducedjoin) == "batting_avg"] <- "BA"
names(reducedjoin)[names(reducedjoin) == "hit_diff"] <- "xHits - H"
names(reducedjoin)[names(reducedjoin) == "hitDiffperAB"] <- "(xHits - H)/AB"
names(reducedjoin)[names(reducedjoin) == "normDiffperAB"] <- "Normalized"
names(reducedjoin)[names(reducedjoin) == "norm_hits_x_ba"] <- "Normalized Hits x BA"

reducedjoin$xBA <- reducedjoin$xHits / reducedjoin$AB
reducedjoin$BAdiff <- reducedjoin$xBA - reducedjoin$BA

```



